maxent model avec xvalidation (year-based test training)

### 2019-04-23 src/maxent_alldata_avec_xvalidation.Rmd
trouble with running for all models.. java issue :/

```{r pre-install & load packages, include=FALSE}
options(java.parameters = "-Xmx1g" )#needs to run before dismo loaded - boosts memory for java to 1g
packages <- c("dismo", "stringr", "rJava", "raster", "ecospat")
source("../src/ipak.R")
ipak(packages)
```

@@

```{r remove unnecessary columns}
presback <- subset(presback, select = -c(cell_id, id, decimalLatitude, decimalLongitude, datecollected, institutioncode, individualcount, depth, resname, originalscientificname, collectioncode, day, nafo_zone, gear, longitude_meters, latitude_meters, total_cell_obs_xy, total_cell_obs_xyt, bottom_depth, XXtotal_cell_obs_xyzt, temp_celsius_depth, temp_celsius_surface, longitude_meters.1, latitude_meters.1, bottom_depth_glorys, longitude_meters.2, latitude_meters.2, cell_id_3d, total_cell_obs_xyzt, cell_id_xyzt))
colnames(presback)
```

@@

```{r}
janvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_winter")
marvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
aprvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
mayvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
junvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
julvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
augvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
sepvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
octvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
novvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
decvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")

mthnavar_lst <- list(jan = janvar, mar = marvar, apr = aprvar, may = mayvar, jun = junvar, jul = julvar, aug = augvar, sep = sepvar, oct = octvar, nov = novvar, dec = decvar)
```

select month
```{r month of interest -change}
mthna <- "jan"
mthno <- "1"
```

```{r subsest data by month}
prback <- subset(presback, month == mthno)
prback <- subset(prback, select = mthnavar_lst[[mthna]])
```

@@

```{r remove na vals}
prback <- na.omit(prback)
```

@@

```{r train and test inc AIC for reg val}
modelmonth <- cbind(paste0(mthna ,unique(prback$year)))
aicetc <- read.csv("../data/env/AICcCalculations.csv", header = TRUE) #pre-created csv file
betavallist <- cbind(aicetc$regularization)

for (year in 1:nrow(modelmonth)){
  timeslice <- modelmonth[year] #testing timeslice
  yr <- str_sub(timeslice, -4) #the year being used for testing
  train <- subset(prback, year != yr) #training data is all years but testing
  trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
  test <-  subset(prback, year == yr) #creates the testing dataset
  testp <- subset(test, occurrence == "1") #splits the testing dataset into presences
  testp <- subset(testp, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  testb <- subset(test, occurrence == "0") #splits the testing dataset into background
  testb <- subset(testb, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling

  for (h in 1:nrow(betavallist)){
    betamval <-  betavallist[h]
    mod <- maxent(x = train, p = trainpa, path = paste0("../output/maxent/models/", timeslice), args=c("maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", paste0("betamultiplier=", betavallist[h])))

@@

    #count number of non-zero coefficients, excluding the final four rows
    coefile <- read.csv(paste0("../output/maxent/models/", timeslice, "/species.lambdas"), header = FALSE, stringsAsFactors=FALSE)
    coefile <- head(coefile, -4)
    coefile2 <- subset(coefile, V2 > 0 | V2 < 0)
    nocoeff <- nrow(coefile2)
    bick <- sum(as.numeric(coefile2[, 2])) #dodgy 
    bic <- (bick*log(sum(trainpa == 1)) - 2*11) #dodgy

@@

    #calculate minimal predicted area using ecospat
    aicetc$mpa_0.9_cum <- ecospat.mpa(samplepred$Cumulative.prediction, perc = 0.9) #90%
      }
    }

@@

@@

for (year in 1:nrow(modelmonth)){
  timeslice <- modelmonth[year] #testing timeslice
  yr <- str_sub(timeslice, -4) #the year being used for testing
  train <- subset(prback, year != yr) #training data is all years but testing
  trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
  test <-  subset(prback, year == yr) #creates the testing dataset
  testp <- subset(test, occurrence == "1") #splits the testing dataset into presences
  testp <- subset(testp, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  testb <- subset(test, occurrence == "0") #splits the testing dataset into background
  testb <- subset(testb, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling

@@

    #calculate minimal predicted area using ecospat
    aicetc$mpa_0.9_cum <- ecospat.mpa(samplepred$Cumulative.prediction, perc = 0.9) #90%

@@

    }
}

@@

```{r binary map using maxsumsenspec}
cumasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^cumu_.*.asc", full.names = TRUE)

>>>>

```{r pre-install & load packages, include=FALSE}
options (java.parameters = "-Xmx2048m")
packages <- c("rJava", "dismo", "stringr", "raster", "ecospat")
source("../src/ipak.R")
ipak(packages)
```

@@

```{r remove unnecessary columns}
presback <- subset(presback, select = -c(cell_id, id, decimalLatitude, decimalLongitude, datecollected, institutioncode, individualcount, depth, resname, originalscientificname, collectioncode, day, nafo_zone, gear, longitude_meters, latitude_meters, total_cell_obs_xy, total_cell_obs_xyt, bottom_depth, XXtotal_cell_obs_xyzt, temp_celsius_depth, temp_celsius_surface, longitude_meters.1, latitude_meters.1, bottom_depth_glorys, longitude_meters.2, latitude_meters.2, cell_id_3d, total_cell_obs_xyzt, cell_id_xyzt, month, depthlayerno, depth_layer))
colnames(presback)
```

@@

```{r}
allvar <- c("occurrence", "year", "temp_surface", "temp_depth", "salinity_surface", "salinity_depth", "o2_depth", "chl_surface", "mlp_surface",  "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
presback <- subset(presback, select = c(allvar))
colnames(presback)
```

@@

```{r remove na vals}
prback <- na.omit(presback)
```

```{r}
train <- prback
trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling
mod <- maxent(x = train, p = trainpa, args=c("maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", "betamultiplier=0.25"))
```

@@

```{r train and test inc AIC for reg val}
aicetc <- read.csv("../data/env/AICcCalculations.csv", header = TRUE) #pre-created csv file
#betavallist <- cbind(aicetc$regularization)
betavallist <- 0.25
  timeslice <- "all_data"
  train <- prback
  trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
  train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling

  #for (h in 1:nrow(betavallist)){
   # betamval <-  betavallist[h]
    mod <- maxent(x = train, p = trainpa, path = paste0("../output/maxent/", timeslice, "/models"), args=c("maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", paste0("betamultiplier=", betavallist)))

@@

    #count number of non-zero coefficients, excluding the final four rows
    coefile <- read.csv(paste0("../output/maxent/", timeslice, "/models/species.lambdas"), header = FALSE, stringsAsFactors=FALSE)
    coefile <- head(coefile, -4)
    coefile2 <- subset(coefile, V2 > 0 | V2 < 0)
    nocoeff <- nrow(coefile2)
    bick <- sum(as.numeric(coefile2[, 2])) #dodgy 
    bic <- (bick*log(sum(trainpa == 1)) - 2*11) #dodgy

    #no presences used in model (less those used for testing)
    maxresults <- read.csv(paste0("../output/maxent/", timeslice, "/models/maxentResults.csv"), header = TRUE)
    notrainingpoints <- maxresults$X.Training.samples
    testauc <- maxresults$Test.AUC
    trainauc <- maxresults$Training.AUC

    #natural log
    samplepred <- read.csv(paste0("../output/maxent/", timeslice, "/models/species_samplePredictions.csv"), header = TRUE)
    samplepred$nat_log <- log(samplepred$Raw.prediction)
    sumsamplepred <- sum(samplepred$nat_log)

@@

    #calculate minimal predicted area using ecospat
    aicetc$mpa_0.9_cum <- ecospat.mpa(samplepred$Cumulative.prediction, perc = 0.9) #90%
      #}

@@

write.csv(aicetc, paste0("../output/maxent/", timeslice, "/models/", betaval, "_AICforMaxentReg.csv"), row.names = FALSE)

@@

#for (year in 1:nrow(modelmonth)){
 # timeslice <- modelmonth[year] #testing timeslice
  #yr <- str_sub(timeslice, -4) #the year being used for testing
  train <- subset(prback, year != yr) #training data is all years but testing
  trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
  test <-  subset(prback, year == yr) #creates the testing dataset
  testp <- subset(test, occurrence == "1") #splits the testing dataset into presences
  testp <- subset(testp, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  testb <- subset(test, occurrence == "0") #splits the testing dataset into background
  testb <- subset(testb, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling

@@

    #calculate minimal predicted area using ecospat
    aicetc$mpa_0.9_cum <- ecospat.mpa(samplepred$Cumulative.prediction, perc = 0.9) #90%
    aicetc$mpa_0.9_clog <- ecospat.mpa(samplepred$Cloglog.prediction, perc = 0.9) #90%

@@

    }
#}

@@

```{r binary maps}
thresholds <- read.csv(paste0("../output/maxent/models/", mthna, "avgAIC.csv"), header = TRUE)
thresholds <- tail(thresholds, 1)
mpa <- thresholds$mpa_0.9_cum
maxsesp <- thresholds$test_maxsumsenspec
cumasc_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^cumu_.*.asc", full.names = TRUE) 
depthlayers <- data.frame(cbind(unique(stdep$ascdepths)))
names(depthlayers)[names(depthlayers) == "cbind.unique.stdep.ascdepths.."] <- "depth"
depthlayers$depth <- as.character(depthlayers$depth)

for (d in 1:nrow(depthlayers)){
  dep <- depthlayers$depth[d]
  for (a in 1:length(cumasc_list)){
    asc <- cumasc_list[a]
    if (grepl(dep, asc)){
      depthlayers[d , paste0("a", colno)] <- asc
        colno <- colno+1
    }
  }
}
write.csv(depthlayers, (paste0("../output/maxent/predictions/", mthna, "/cumu_asclayerbydepth.csv")), row.names = TRUE)

for (x in 1:nrow(depthlayers)){
  asclay <- depthlayers[x, ]
  asclay <- asclay[!is.na(asclay)]
  asclay <- as.data.frame(rbind(asclay))
  asclay$V1 <- as.character(asclay$V1) #v1 corresponds to depth
  asclay$V2 <- as.character(asclay$V2) #v2 to end corresponds to year
  asclay$V3 <- as.character(asclay$V3)
  asclay$V4 <- as.character(asclay$V4)
  asclay$V5 <- as.character(asclay$V5)

  asc2003 <- raster(asclay$V2)
  asc2004 <- raster(asclay$V3)
  asc2005 <- raster(asclay$V4)
  asc2006 <- raster(asclay$V5)

  stkasc <- stack(asc2003, asc2004, asc2005, asc2006)
  avg <- calc(stkasc, fun = mean)

  writeRaster(avg, (paste0("../output/maxent/predictions/", mthna, "/cumu_avg_", asclaydep, ".asc")), overwrite=TRUE, args='outputformat=cumulative')

cumavg_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^cumu_avg_.*.asc", full.names = TRUE) 
for(lyr in 1:length(cumavg_list)){
  avgras <- raster(cumavg_list[lyr])
  reclassmpa <- c(mpa, Inf, 1)
  reclassmpa <- matrix(reclassmpa, ncol = 3, byrow = TRUE)
  binavgras <- reclassify(avgras, reclassmpa)
}
